---
title: 'Getting data into R'
output:
  html_document:
    highlight: tango
    theme: yeti
    toc: yes
    toc_float: yes
---

----- 

# Introduction
In this lab we will introduce ways to get data into R, and how to start exploring the data. 


### Student Learning Objectives
After completing this lesson students will be able to

* Import data that is stored in an external Excel, CSV or tab-delimited text file into R.
* Identify missing data elements in a data table. 
* Identify data types for each variable in a data frame. 

### What's a delimiter? 
Before we go much further, let's talk about the word **_delimiter_**. This is the character, or symbol, that separates columns of data from each other when stored in an external file. This file could be a simple `.txt` text file or an `.xlsx` Excel file. Recall back to the Data Camp exercises on data frames and matrices. Each column in the matrix represented data on a specific variable. Something had to tell R how to distinguish which values went with which variable. 

There are two main types of delimiters we will consider in this class; comma and tab. That does not mean that data can't be stored in other ways, these are just the two most common. 


### What's a path? 

The _path_ is a programmatic way to direct your computer to the location of a file. It's like the file's address, or, where it lives on your computer. 

In all the examples below, the path shown (`data/`) is the path on **MY** computer.You will have to update this path to **YOUR** path on **YOUR** machine that points to your MATH130 folder. 

Revisit the [Software Overview](https://norcalbiostat.netlify.com/post/software-overview/) blog post from Dr. Donatello's website on how to find your path. 

### Download data
* From the [Data](https://norcalbiostat.netlify.com/data/cleaned_data/) page download the following data sets **into your MATH130 folder**: `NCBirths`, `email`, and `Police Shootings`. 
* **DO NOT RENAME ANYTHING** If you download multiple copies of the data, do not keep the versions that have a (1) or some number in the name. 



----- 

# Importing Data

In this lesson we are only going to explore reading flat files that exist on your computer into R from three most commonly used data sources: A tab-delimited text file, A CSV file and an Excel file. Ideally each row represents data on a single observation and each column contains data on a single variable, or characteristic, of the observation. This is called `tidy data`. We will start to learn some tools to look at the data, and for getting data from an external file into R for analysis. 

#### Look at your data!!

The first thing you should always do is look at your data and ask yourself the folliwng things: 

1. Were the variable names read in correctly? 
2. Is there the expected number of rows and columns? 
3. Are the data types for each variable as expected? 

There are many ways this can be done, here are three

1. `str()` to show you the structure of the data frame. 
2. `head()` to show you the top 6 rows
3. `data[1:x, 1:x]` - when the data frame is too large for the above two to work well, we can use matrix notation to view only the first X columns, and first X rows. 


### Text files
Text files are very simple files that have a `.txt` file extension. Common delimiters include a space, a comma (,) or a tab. Uncommon delimiters could include a % or even a semi-colon. Follow along with these examples and make sure you can read in the data correctly and that it matches what is shown below. 

We will use the `read.table()` function that is in base R to read in any type of delimited file. A tab-delimited text file can be read in using `"\t"` as the deliminator character. In this class you **ALWAYS** want to include `header=TRUE` to signify that the data in the first row contains our column names. 
```{r}
email <- read.table("data/email.txt", header=TRUE, sep="\t")
```


```{r}
str(email)
```

* The `email` data set has 3921 observations, and 21 variables. 
* The variable names were read in correctly, as evidenced by the words in the far right corner after the `$`. 
* Most variables are treated as numeric integers, except a few are factors. Including `time`. If we wanted to examine a time trend, we would need to convert the class into a `datetime` 


### Excel files
The best method I have found so far to read in Excel files is from the [`readxl` package](http://readxl.tidyverse.org/) by [Hadley Wickham](http://hadley.nz/). This package need to be installed first, and then can be simply loaded each time you start an `R` session where you will be reading in this type of data. Go ahead and install it now.  

The `read_excel()` function is what we are going to use. Note the use of the underscore `_` instead of a period `.` between read and the file type.  

```{r}
library(readxl)
police  <- read_excel("data/WashPost Police Shootings 2015.xlsx", sheet=1, col_names=TRUE)
police[1:10,1:5]
```

* The variable `date` is a `dttm` or _date-time_ variable. This means `R` recognizes it directly as a date, not some string of numbers. 
* Categorial variables such as `manner_of_death` and `city` are read in as `character` instead of factor. We may or may not want to change that later.
* Numeric varibles such as `age` are of type `dbl` (double). This is similar to `integer` or `numeric`, so we are fine. 


### CSV: Comma Separated Values
CSV is a fancy way of saying a text file with comma-separated values (i.e. CSV). This file type can also open in Excel, the icons even look similar. 

We could use `read.table()` to import this data file, but `read.csv()` is optimized to read in CSV files. 

```{r}
NCbirths <- read.csv("data/NCbirths.csv", header=TRUE)
head(NCbirths)
```

* The `NCbirths` data set has 1000 rows and 13 columns. 
* Variable names are read in OK. 
* Most the data is beign read in as factors and integers. 
* There are some `NA` values for the variable `fage` (fathers age). 

----- 

# Missing Data

* R puts a `NA` as a placeholder when the value for that piece of data is missing. 
* Missing data can be a result of many things: skip patterns in a survey (i.e. non-smokers don't get asked how many packs per week they smoke), errors in data reads from a machine, etc. 


**Problem 1** 
`R` can't do arithmetic on missing data. So `5 + NA = NA`, and if you were to try to calculate the `mean()` of a variable, you'd also get `NA`. 

```{r}
mean(NCbirths$fage)
```

**Problem 2**
Some plots will show `NA` as it's own category, which is undesirable. 

```{r, echo=FALSE}
library(ggplot2)
ggplot(NCbirths, aes(premie))+geom_bar()
```


**How to identify if your data has missing values**

1. Look at the raw data using `head()` or `str()`
2. Look at data summaries using `table()` for categorical data and `summary()` for numerical data. 

```{r}
table(NCbirths$habit, useNA="always")
summary(NCbirths$fage)
```

Later labs will discuss how to work around these missing values.

----- 

# On Your Own

1. Right click this link [Lab 2.Rmd](lab2.Rmd) link and **Save As** (or Save Link As) to download the RMarkdown file. 
    - Save this in your class folder. 
    - Rename this file using your username `lab2_username`. 
2. This is another homework template for you to use to write your assignment. Write your answers directly into this document. 
    - Don't forget to knit your document early and often!
3. Submit the compiled **PDF** or **WORD** file by the due date.


----- 

# Additional resources

### Importing Data

* https://www.datacamp.com/courses/importing-data-into-r

### Missing Data

* http://www.statmethods.net/input/missingdata.html
* https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/
* http://faculty.nps.edu/sebuttre/home/R/missings.html


