---
title: 'Lesson 05 - Controlling your factors'
author: "Robin Donatello"
date: "Last Updated `r format(Sys.Date(), '%m-%d-%Y')`"
output:
  html_document:
    highlight: tango
    theme: yeti
    toc: yes
    toc_float: yes
---

----- 

# Introduction
In this lesson we will discuss ways to organize and deal with categorical data, also known as factor data types. 


### Student Learning Objectives
After completing this lesson students will be able to

* Create and control the labeling and ordering of categorical variables (factors).

We'll use the data set on emails for this lesson. 

```{r}
email <- read.table("data/email.txt", header=TRUE, sep="\t")
```

----- 

# What is a factor?
The term factor refers to a statistical data type used to store categorical variables. The difference between a categorical variable and a continuous variable is that a categorical variable corresponds to a limited number of categories, while a continuous variable can correspond to an infinite number of values.

An example of a categorical variable is the `number` variable in the `email` data set. This variable contains data on whether there was no number, a small number (under 1 million), or a big number in the content of the email. 

```{r}
table(email$number)
class(email$number)
```

We can see that `R` recognizes this variable as a factor variable, and that it has the three levels: `big`, `none`, and `small`. 

```{r}
levels(email$number)
```

# Creating factor variables

Typically data are entered into the computer using numeric codes such as 0 and 1. These codes stand for categories, such as "no" and "yes". Sometimes we want to analyze these binary variables in two ways: 

* For statistical analyses, the data must be numeric 0/1.
* For many graphics, the data must be a factor, "no/yes".

**Example: Is the email flagged as spam?**
The `spam` variable is recorded as an integer variable with values 0 and 1. 
```{r}
table(email$spam)
class(email$spam)
```

We use the function `factor()` to convert the numeric variable `spam` to a factor, applying `lables` to convert 0 to "no" and 1 to "yes". 

```{r}
email$spam_fac <- factor(email$spam, labels=c("no", "yes"))
```

The ordering of the `labels` argument _must_ be in the same order (left to right) as the factor levels themselves. Look back at the order of columns in the `table` - it goes 0 then 1. Thus our labels need to go "no" then "yes". 

**Always confirm your recode** 

Here we confirm that the new variable was created corectly by creating a two-way contingency table by calling the `table(old variable, new variable)` function on both the old and new variables. 

```{r}
table(email$spam, email$spam_fac, useNA="always")
```

Here we see that all the 0's were recoded to 'no's, and all the 1's recoded to "yes"'s, and there are no new missing values. Success!



# Ordering of Factors
Let's revisit the variable `number`, that contains the size of the number in the email. 

```{r}
table(email$number)
```

Specifically the ordering from left to right of the factors. Seem odd? This is ordinal data, in that `none` is inherently "smaller" than `small`, which is smaller than `big`. But the ordering goes `big` - `none` - `small`, which is in Alphabetical order!

Gee thanks R. Exactly how I want my factors ordered (NOT!)

We need to take control of these factors! We can do that by re-factoring the existing factor variable, but this time specifying the `levels` of the factor (since it already has labels).

```{r}
email$number <- factor(email$number, levels=c("none", "small", "big"))
table(email$number)
```

There we go. Now it's in a readable, left to right in increasing size order. This will be important for graphing. 


----- 

# Additional resources

* STAT 133 UC Berkeley https://www.stat.berkeley.edu/classes/s133/factors.html
* Be the boss of your factors using `dplyr` and `forcats`  http://stat545.com/block029_factors.html

