---
title: "Lab 4"
author: "NAME"
date: "DATE"
output: html_document
---

# Introduction

This lab uses the `NCbirths` data set. Use the following code chunk to read the data into R. Need help? Look at lesson 03 - importing data.

```{r}

```

You will use a combination of base and `ggplot2` graphics for this lab. Your answers can be for your eyes only (exploratory). They do not have to be pretty.  

The code chunk below loads the `ggplot2` package for you, and also sets some code chunk options (using `opts_chunk` from the `knitr` package) to make your knitted report output more readable. I encourage you to play around with these options to learn how they work. 

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(knitr)
opts_chunk$set(warning=FALSE, message=FALSE, fig.height=4, fig.width=5, fig.align='center') 
```

# Univariate

1. A table of marital status (`marital`)


2.  A barchart of maritial status


3. A histogram of mothers age (`mage`)

    a.  Now add an overlaid density plot in a different color. Don't forget to adjust the `aes` on the histogram.

4. A horizontal boxplot of weight of the baby (`weight`)



# Bivariate

5. Create a two-way frequency table of maturity status (`mature`) against smoking `habit`


6. Create a proportion table of smoking habit _within_ maturity status. Round to 3 digits. 


7. Create a grouped barchart that reflects the proportions you calculated above. Think carefully which variable goes on the `x` axis, and which one is used for the `fill`


8. A scatterplot of length of pregnancy in `weeks` and the babies `weight`. 
    a. Include a smoother line in brown, and a best fit linear model line in purple
    

9. Grouped boxplots of baby `weight` by mothers smoking `habit`. Make sure you `fill` the boxes by `habit` as well. 


10. Replicate the same plot as above, but overlay a violin plot and change the transparency of both violin and boxplot layers. 


11. Overlaid density plots of weight `gained` by babies `gender`. No fill, color only.  



# Bonus optional questions. 
Looking for more? Want to start learning how to clean up your plots to make them professional looking? Go check out the [full data viz tutorial](https://norcalbiostat.github.io/MATH130/full_data_viz_tutorial.html) and try the following problems. 

1. A scatterplot of length of pregnancy in `weeks` and the babies `weight`, color the points and lowess lines by `gender`. 


2. Use `facet_wrap` to create side by side boxplots with overlaid violins (using good trasparency options) of `weight` by `mature`, paneled by `whitemom`. 

3. Remove the NA group from the previous plot and recreate the plot. 



4. Now add points for the means to the above plot. 


5. Use `coord_flip()` to make the boxes horizontal



6. Go to the [R Cookbook](http://www.cookbook-r.com/Graphs/) and learn how to remove that legend. 

7. Apply a `theme_minimal()` or `theme_bw()` layer to remove the grey background from the plot. 
