---
title: "Lab 4 - Data Visualization"
author: "Solutions"
date: "Last Updated `r format(Sys.Date(), '%m-%d-%Y')`"
output:
  pdf_document: default
  html_document: default
---

# Introduction

This lab uses the `NCbirths` data set. Use the following code chunk to read the data into R. Need help? Look at lesson 03 - importing data.

```{r}
NCbirths <- read.csv("data/NCbirths.csv", header=TRUE)
```

You will use a combination of base and `ggplot2` graphics for this lab. Your answers can be for your eyes only (exploratory). They do not have to be pretty.  

The code chunk below loads the `ggplot2` package for you, and also sets some code chunk options (using `opts_chunk` from the `knitr` package) to make your knitted report output more readable. I encourage you to play around with these options to learn how they work. 

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(knitr)
opts_chunk$set(warning=FALSE, message=FALSE, fig.height=4, fig.width=5, fig.align='center') 
```

# Univariate

1. A table of marital status (`marital`)

```{r}
table(NCbirths$marital)
```

2.  A barchart of maritial status

```{r}
ggplot(NCbirths, aes(x=marital)) + geom_bar()
```


3. A histogram of mothers age (`mage`)

```{r}
ggplot(NCbirths, aes(x=mage)) + geom_histogram()
```

    a.  Now add an overlaid density plot in a different color. Don't forget to adjust the `aes` on the histogram.
```{r}
ggplot(NCbirths, aes(x=mage)) + geom_density(col="blue") + 
  geom_histogram(aes(y=..density..), colour="black", fill=NA)
```
    
4. A horizontal boxplot of weight of the baby (`weight`)

```{r}
boxplot(NCbirths$weight, horizontal=TRUE)
```


# Bivariate

5. Create a two-way frequency table of maturity status (`mature`) against smoking `habit`

```{r}
table(NCbirths$mature, NCbirths$habit)
```

6. Create a proportion table of smoking habit _within_ maturity status. Round to 3 digits. 

```{r}
round(prop.table(table(NCbirths$mature, NCbirths$habit), margin=1),3)
```

7. Create a grouped barchart that reflects the proportions you calculated above. Think carefully which variable goes on the `x` axis, and which one is used for the `fill`

```{r}
ggplot(NCbirths, aes(x=mature, fill=habit)) + geom_bar(position="dodge")
```

8. A scatterplot of length of pregnancy in `weeks` and the babies `weight`. 
    a. Include a smoother line in brown, and a best fit linear model line in purple
    
```{r}
ggplot(NCbirths, aes(x=weeks, y=weight)) + geom_point() + 
  geom_smooth(se=FALSE, method="lm", color="blue") + 
  geom_smooth(se=FALSE, color="red")
```

9. Grouped boxplots of baby `weight` by mothers smoking `habit`. Make sure you `fill` the boxes by `habit` as well. 

```{r}
ggplot(NCbirths, aes(x=habit, y=weight, fill=habit)) + geom_boxplot()
```

10. Replicate the same plot as above, but overlay a violin plot and change the transparency of both violin and boxplot layers. 

```{r}
ggplot(NCbirths, aes(x=habit, y=weight, fill=habit)) + 
        geom_boxplot(alpha=.6) + geom_violin(alpha=.2)
```

11. Overlaid density plots of weight `gained` by babies `gender`. No fill, color only.  

```{r}
ggplot(NCbirths, aes(x=gained, col=gender)) + geom_density()
```


# Bonus optional questions. 
Looking for more? Want to start learning how to clean up your plots to make them professional looking? Go check out the [full data viz tutorial](https://norcalbiostat.github.io/MATH130/full_data_viz_tutorial.html) and try the following problems. 

1. A scatterplot of length of pregnancy in `weeks` and the babies `weight`, color the points and lowess lines by `gender`. 
```{r}
ggplot(NCbirths, aes(x=weeks, y=weight, col=gender)) + geom_point() + geom_smooth(se=FALSE)
```


2. Use `facet_wrap` to create side by side boxplots with overlaid violins (using good trasparency options) of `weight` by `mature`, paneled by `whitemom`. 
```{r}
ggplot(NCbirths, aes(x=mature, y=weight, fill=mature)) + 
        geom_boxplot(alpha=.6) + geom_violin(alpha=.2) + facet_wrap(~whitemom)
```

3. Remove the NA group from the previous plot and recreate the plot. 

```{r}
library(dplyr)
plot.data <- NCbirths %>% select(mature, weight, whitemom) %>% na.omit()
ggplot(plot.data, aes(x=mature, y=weight, fill=mature)) + 
        geom_boxplot(alpha=.6) + geom_violin(alpha=.2) + facet_wrap(~whitemom)
```


4. Now add points for the means to the above plot. 
```{r}
ggplot(plot.data, aes(x=mature, y=weight, fill=mature)) + 
        geom_boxplot(alpha=.6) + geom_violin(alpha=.2) + facet_wrap(~whitemom) + 
        stat_summary(fun.y="mean", geom="point", size=3, pch=17,color="blue")
```


5. Use `coord_flip()` to make the boxes horizontal

```{r}
ggplot(plot.data, aes(x=mature, y=weight, fill=mature)) + 
        geom_boxplot(alpha=.6) + geom_violin(alpha=.2) + facet_wrap(~whitemom) + 
        coord_flip()
```


6. Go to the [R Cookbook](http://www.cookbook-r.com/Graphs/) and learn how to remove that legend. 
```{r}
ggplot(plot.data, aes(x=mature, y=weight, fill=mature)) + 
        geom_boxplot(alpha=.6) + geom_violin(alpha=.2) + facet_wrap(~whitemom) + 
        coord_flip() + scale_fill_discrete(guide=FALSE)
```

7. Apply a `theme_minimal()` or `theme_bw()` layer to remove the grey background from the plot. 
```{r}
ggplot(plot.data, aes(x=mature, y=weight, fill=mature)) + 
        geom_boxplot(alpha=.6) + geom_violin(alpha=.2) + facet_wrap(~whitemom) + 
        coord_flip() + scale_fill_discrete(guide=FALSE) + theme_bw()
```
