---
title: 'Lab 2: Importing, exploring, managing data using functions'
author: "Solutions"
date: "DATE"
output: pdf_document
---

Answer the questions in this lab and submit the compiled WORD or PDF by the deadline. 

## Importing Data

1. When reading in the `Police Shootings` Excel data set, what do the arguments `sheet=1` and `col_names=TRUE` mean? 

> `sheet=1` Takes the data from the first sheet. 
> `col_names` use the first row as column names. 


Use the `NCbirths` data set to answer the next set of questions. Read in the data set in the code chunk below. 

```{r}
nc <- read.csv("data/NCbirths.csv", header=TRUE)
```


2. How many observations and variables are contained in this data set? 

```{r}
dim(nc)
```

> There are 1000 observations and 13 variables

* they don't have to have used `dim`. they could have looked in the environment. 

3. Calculate the mean age of the mothers (`mage`) in the sample.
```{r}
mean(nc$mage)
```


4. Pregnancies last on average 38 weeks. Edit the `weeks` variable to change all records where `weeks` is greater than 38, to equal 38. That is, for all record where `weeks>38`, change the value of `weeks` to `<- 38`. 

```{r}
nc$weeks[nc$weeks>38] <- 38
max(nc$weeks, na.rm=TRUE) 
```

They need to have confirmed that they did the recoding correctly. 


5. Use the `summary` function to calculate summary statistics on the fathers age (`fage`). 
   Round to 3 digits using the `digits=` argument. Don't forget that you can look at the 
   bottom of the help for summary (`?summary`) file for examples on how to use this function.  
```{r}
summary(nc$fage, digits=3)
```

   
6. What is the distribution of smoking habit (`habit`) amongst the mothers in the sample? 
   i.e. How many are smokers and how many are non-smokers?  _Hint: Use the `table()`_ function. 
```{r}
table(nc$habit)
```
   
> There are 873 non-smokers and 126 smokers in the sample.  
   
7. Use the `ifelse()` function to create a new variable called `missing_fage` to identify if the fathers age is missing. The **logical statement** to identify if something is missing looks like this: `is.na(variable)`. 
    * Set this new variable equal to `'MISSING'` if `fage` is missing (the logical statement is TRUE)
    * set this new variable equal to `'OBSERVED'` if `fage` is not missing (the logical statement is FALSE)

```{r}
nc$missing_fage <- ifelse(is.na(nc$fage), 'MISSING', 'OBSERVED')
table(nc$missing_fage)
```

* They have to make this variable on the nc data set. 
* They also have to do something to show that it was created. 

8. What class of data is this new variable? 
```{r}
class(nc$missing_fage)
```

> This new variable is a character variable.     

9. What percent of records are missing the fathers age? 

```{r}
mean(is.na(nc$fage))*100
```

> 17.1% of the data on the fathers age is missing. 

* There are many ways they can get this answer. 





