---
title: "Lab 2: Importing, exploring, managing data using functions"
author: "NAME"
date: "DATE"
---

Answer the questions in this lab and submit the compiled WORD or PDF by the deadline. 

## Importing Data

1. When reading in the `Police Shootings` Excel data set, what do the arguments `sheet=1` and `col_names=TRUE` mean? 


Use the `NCbirths` data set to answer the next set of questions. Read in the data set in the code chunk below. 

```{r}
# Read in the NC Births data set here, then delete this comment
```


2. How many observations and variables are contained in this data set? 


3. Calculate the mean age of the mothers (`mage`) in the sample.
```{r}

```


4. Pregnancies last on average 38 weeks. Edit the `weeks` variable to change all records where `weeks` is greater than 38, to equal 38. That is, for all record where `weeks>38`, change the value of `weeks` to `<- 38`. 

```{r}

```


5. Use the `summary` function to calculate summary statistics on the fathers age (`fage`). 
   Round to 3 digits using the `digits=` argument. Don't forget that you can look at the 
   bottom of the help for summary (`?summary`) file for examples on how to use this function.  
```{r}

```

   
6. What is the distribution of smoking habit (`habit`) amongst the mothers in the sample? 
   i.e. How many are smokers and how many are non-smokers?  _Hint: Use the `table()`_ function. 
```{r}

```
   
   
7. Use the `ifelse()` function to create a new variable called `missing_fage` to identify if the fathers age is missing. The **logical statement** to identify if something is missing looks like this: `is.na(variable)`. 
    * Set this new variable equal to `'MISSING'` if `fage` is missing (the logical statement is TRUE)
    * set this new variable equal to `'OBSERVED'` if `fage` is not missing (the logical statement is FALSE)

```{r}

```



8. What class of data is this new variable? 
```{r}

```
   
 
 
   
9. What percent of records are missing the fathers age? 






