---
title: "Lab 3: More data management and grouping using factors"
author: "NAME"
date: "DATE"
---

Answer the questions in this lab and submit the compiled HTML or PDF by the deadline. Don't forget to change your name and date in the above header. 

This lab uses the `dplyr` and `nycflights13` packages. It is good habit to load all packages in the first code chunk. 

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(nycflights13)
```


# Basic verbs
You will use the `flights` data set for the next few exercises.  Let's load that into our working environment so we can look at it.
```{r}
flights <- flights
```

* At each step use the assignment operator `<-` to store the results into a new data table and use that data in the next step. 
* At each step, print out the resulting data frame so you can see the results. 

1. Use `select()` to extract the following variables: `origin`, `distance`, and `air_time`, `dest`. 

```{r}
f1 <- flights %>% select(origin, distance, air_time, dest)
f1
```

2. Use `filter()` to select only the flights whose destination (`dest`) is Atlanta (ATL)

```{r}
f2 <- f1 %>% filter(dest == 'ATL')
f2
```

3. Use `mutate()` to create a new variable `speed` that calculates speed of the plane as `distance/air_time*60.

```{r}
f3 <- f2 %>% mutate(speed = distance/air_time*60)
f3
```


# How many passengers can a plane hold before needing another engine? 
This question uses the `planes` data set. Let's load that into our working environment so we can look at it.
```{r}
planes <- planes
```


1. Examine the variable `engines` using `table()` and `class`. What is it's data type? 


```{r}
table(planes$engines)
class(planes$engines)
```

The number of engines is an integer variable with values between 1 and 4. 

2. There are too few planes with more than 2 engines. Recode all records with 4 engines to a value of 3. _Hint: Revisit lesson 04_. Create a `table` of this variable again to ensure that all 4's are now 3's. 

```{r}
planes$engines[planes$engines==4] <- 3
table(planes$engines)
```


3. Create a new factor variable `num_engines` from `engines` with labels "one", "two", "three+". 

```{r}
planes$num_engines <- factor(planes$engines, labels=c("one", "two", "three+"))
```


4. Create a two-way `table` of `engines` against `new_engines` to confirm that this new factor variable was created correctly. 

```{r}
table(planes$num_engines, planes$engines)
```


5. Use `dplyr` chaining magic to...

* take the planes data set _and then_...
* `group_by`the `num_engines` _and then_...
* use `summarise` to create three new variables: 
    - `ave_seats` as the `mean()` number of `seats`
    - `min_seats` as the `min()` number of `seats`
    - `max_seats` as the `max()` number of `seats`


```{r}
planes %>% group_by(num_engines) %>% summarise(ave_seats = mean(seats), 
                                               min_seats = min(seats), 
                                               max_seats = max(seats))
```

