---
title: "Assignment 1"
author: "STUDENT NAME"
date: "DATE"
output: html_document
linkcolor: blue
---

Here's your chance to demonstrate that you can integrate the topics and skills you learned so far into a literate report. 

When the question asks you to perform a coding task, insert a code chunk after each question where you will write the code to answer that question. Knitting after each completed code chunk will help you to ensure your final product works as intended! That way if it breaks, you know exactly where the error lies. It's like saving after every answer!

The first question is done for you as an example. 

# Object assignments

0. Calculate 3 + 4. Put the answer in the grey area below and knit the document. Make sure you can find this code and output in the resulting HTML file. 
```{r}
3+4
```


1. Calculate $2^5$ by typing this mathematical expression in the code chunk below and then knit the document. 
```{r}

```

2. What are the values after each statement in the following?

```{r}
mass <- 47.5            # a
age  <- 122             # b
mass <- mass * 2.0      # c
age  <- age - 20        # d
mass_index <- mass/age  # e
```

a. `mass` = 
b. `age` = 
c. `mass` = 
d. `age` = 
e. `mass_index` = 


3. Assign a numeric value to the variable `my_apples`, assign a different numeric value to the variable `my_oranges`. Add these two together and assign the result to the variable `my_fruit`. Print the result of `my_fruit` to the report. 
```{r}

```

4. What type of data is `my_fruit`? Use a function discussed in lesson 03 to have R show the answer to this question.
```{r}

```


5. A variable that classified as `logical` can hold only what two values? These are also called Boolean variables. 


6. Knit this document. 


# Vectors

1. The following vectors have been created for you. Use the `class()` function to identify the type of vector for each of these examples. Report your answer for each in a sentence below. 

```{r}
num_char <- c(1, 2, 3, "a")
num_logical <- c(1, 2, 3, TRUE)
char_logical <- c("a", "b", "c", TRUE)
tricky <- c(1, 2, 3, "4")
```


Let's go to Vegas!

1. Create three vectors (`weekday`, `poker`, `roulette`) using the `c()` operator to describe the following outcome. 

* `weekday`: The 5 weekdays. Use these three letter codes: mon, tue, wed, thu, fri.
* `poker`: 
    - On Monday you won $140 
    - Tuesday you lost $50
    - Wednesday you won $20
    - Thursday you lost $120
    - Friday you won $240
* `roulette`:
    - On Monday you lost $24
    - Tuesday you lost $50
    - Wednesday you won $100
    - Thursday you lost $350
    - Friday you won $10


_Hint: If I won $30 on monday and lost $20 on tuesday, my vector would look like `c(30, -20)`_
```{r}

```

2. Use the `sum()` function to calculate how much money did you gain/loose on each game.
```{r}

```

3. Add the results of both vectors together, then calculate your net gain over the weekend. Did you come out ahead? 
```{r}

```



4. On which days did you make money on poker? _Hint: subset the `weekday` vector using a logical statement about `poker`_
```{r}

```




5. On which days did you do better on poker than roulette? _Hint: compare the two vectors using a statement of inequality_

```{r}

```


6. Knit this document to make sure it still works. 


