---
title: "HW 2: Importing, exploring, managing data using functions"
author: "NAME"
date: "DATE"
---

Answer the questions in this lab and submit the compiled WORD or PDF by the deadline. 


1. Questions in this assignment use the `NCbirths` data set. Read in the data set in the code chunk below. 

```{r}
# Read in the NC Births data set here, then delete this comment
```


2. When reading in the `Police Shootings` Excel data set, what do the arguments `sheet=1` and `col_names=TRUE` mean? 



3. Use the `str()` function to determine how many observations and variables are contained in the `NCbirths` data set. 
Write the code necessary, then answer the question in a complete sentence. 



4. Go to Dr. D's data website (http://norcalbiostat.com/data/cleaned_data/) and find 3 data sets with different file types.
For each file state the name of the file, it's file type, and the function you would need to use to read that file into R. 

> Ex: The `email` data set is tab-delimited text (.txt) file. It can be read into R using the `read.delim()` function. 

* 

* 

* 



5. Calculate the mean age of the mothers (`mage`) in the sample.
```{r}

```


6. Pregnancies last on average 38 weeks. Edit the `weeks` variable to change all records where `weeks` is greater than 38, to equal 38. That is, for all record where `weeks>38`, change the value of `weeks` to `<- 38`. 

```{r}

```


7. Use the `summary` function to calculate summary statistics on the fathers age (`fage`). 
   Round to 3 digits using the `digits=` argument. Don't forget that you can look at the 
   bottom of the help for summary (`?summary`) file for examples on how to use this function.  
```{r}

```

   
8. What is the distribution of smoking habit (`habit`) amongst the mothers in the sample? 
   i.e. How many are smokers and how many are non-smokers?  _Hint: Use the `table()`_ function. 
```{r}

```
   
   
9. Use the `ifelse()` function to create a new variable called `missing_fage` to identify if the fathers age is missing. 
The **logical statement** to identify if something is missing looks like this: `is.na(variable)`. 
    * Set this new variable equal to `'MISSING'` if `fage` is missing (the logical statement is TRUE)
    * set this new variable equal to `'OBSERVED'` if `fage` is not missing (the logical statement is FALSE)

```{r}

```



10. What class of data is this new variable? _Hint: Use the `class()` function._
```{r}

```
   
 
 
   
11. What percent of records are missing the fathers age? There are several ways you can calculate this! 
Explain what your process, and answer the question in a full sentence. 






