---
title: "Homework 4"
author: "NAME"
date: "DATE"
output: html_document
---

# Introduction

You will be managing factors and creating some basic plots in this homework assignment. The data sets you will use come from two places, `NCbirths` which you will have to read in below, and the `planes` data set which comes part of the `nycflights13` package. The `planes` data set contains data on plane characteristics such as the number of engines. 

The code chunk below sets some code chunk options (using `opts_chunk` from the `knitr` package) to make your knitted report output more readable. I encourage you to play around with these options to learn how they work. 

```{r, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.height=4, fig.width=5, fig.align='center') 
# Load the ggplot2 library here
# Load the nycflights13 library here
# Read in the NCbirths data set here. 
```

You can use a combination of base and `ggplot2` graphics for this lab. Your answers can be for your eyes only (exploratory). They do not have to be pretty.  

# Data Recoding

> How many engines to planes typically have? 

This question uses the `planes` data set which is built into the `nycflights13` package. Let's load that into our working environment so we can look at it.
```{r}
data(planes)
```


1. Examine the variable `engines` using `table()` and `class`. What is it's data type? 

```{r}

```


2. There are too few planes with more than 2 engines. Recode all records with 4 engines to a value of 3. _Hint: Revisit the earlier lesson on recoding data_. Create a `table` of this variable again to ensure that all 4's are now 3's. 

```{r}

```


3. Create a new factor variable `num_engines` from `engines` with labels "one", "two", "three+". Make sure you create this new variable as part of the `planes` data set. (i.e. `planes$num_engines <-` )

```{r}

```


4. Create a two-way `table` of `engines` against `num_engines` to confirm that this new factor variable was created correctly. 

```{r}

```


# Univariate plots
This section uses data from both the `NCbirths` and the `planes` data sets. 


1. A table of marital status (`marital`) from NCbirths. 



2. A barchart of maritial status (same as above)



3. A barchart of the type of plane (`planes$type`) with the levels displayed in increasing frequency order. 



4. A histogram of mothers age (`mage`) with an overlaid density plot in a different color. Be sure that both density curve and histogram can be seen. 



5. A horizontal boxplot of weight of the baby (`weight`)



# Bivariate plots

6. Create a two-way frequency table of maturity status (`mature`) against smoking `habit`



7. Create a proportion table of smoking habit _within_ maturity status. Round to 3 digits. 



8. Create a grouped barchart that reflects the frequencies you calculated above. Think carefully which variable goes on the `x` axis, and which one is used for the `fill`



9. A scatterplot of length of pregnancy in `weeks` and the babies `weight`. Include a smoother line in brown, and a best fit linear model line in purple
    
    

10. Grouped boxplots of baby `weight` by mothers smoking `habit`. Make sure you `fill` the boxes by `habit` as well. 



11. Replicate the same plot as above, but overlay a violin plot and change the transparency of both violin and boxplot layers. 



12. Overlaid density plots of weight `gained` by babies `gender`. Do not apply a `fill`, only use the `color` aestetic. 



