---
title: "Lesson 07 - Creating graphics"
author: "Robin Donatello"
date: "Last Updated `r format(Sys.Date(), '%m-%d-%Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
---


```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
opts_chunk$set(warning=FALSE, message=FALSE) 
```

> This lesson is a brief excerpt from Dr. Donatello's [[Full data viz tutorial]](full_data_viz_tutorial.html). 

----

# Introduction
Visualizing your data is hands down the most important thing you can learn to do. Seeing is critical to understanding. There are two audiences in mind when creating data visualizations: 

1. For your eyes only (FYEO). These are quick and dirty plots, without annotation. Meant to be looked at once or twice.
2. To share with others. These need to completely stand on their own. Axes labels, titles, colors as needed, possibly captions.

You will see, and slowly learn, how to add these annotations and how to clean up your graphics to make them sharable. Functions inside the `ggplot2` package automatically does a lot of this work for you. Remember this package has to be loaded prior to being able to access the functions within. 

```{r}
library(ggplot2)
```

Use the navbar to the left to help you navigate this long document. 

## The syntax of `ggplot`
The reason we use the functions in `ggplot2` is for consistency in the structure of it's arguments. Here is a bare bones generic plotting function: 

```{r, eval=FALSE}
ggplot(data, aes(x=x, y=y, col=col, fill=fill, group=group)) +  geom_THING() 
```

### Required arguments

* `data`: What data set is this plot using? This is ALWAYS the first argument.
* `aes()`: This is the _aestetics_ of the plot. What's varible is on the x, what is on 
   the y? Do you want to color by another variable, perhaps fill some box by the value
   of another variable, or group by a variable. 
* `geom_THING()`: Every plot has to have a geometry. What is the shape of the thing you 
   want to plot? Do you want to plot points - use `geom_points()`. Want to connect those
   points with a line? Use `geom_lines()`. We will see many varieties in this lab. 


## The Diamonds Data

We will use a subset of the `diamonds` dataset that comes with the `ggplot2` package. This dataset contains the prices and other attributes of almost 54,000 diamonds. Review `?diamonds` to learn about the variables we will be using. 

```{r}
library(ggplot2)
data("diamonds")
set.seed(1410) # Make the sample reproducible
dsmall <- diamonds[sample(nrow(diamonds), 1000), ]
```

----

# Univariate (One Variable)


## Categorical variables
Both Nominal and Ordinal data types can be visualized using the same methods: tables, barcharts and pie charts. 

### Tables
Tables are the most common way to get summary statistics of a categorical variable. The `table()` function produces a frequency table, where each entry represents the number of records in the data set holding the corresponding labeled value. 
```{r}
table(dsmall$cut)
```
There are 27 Fair quality diamonds, 83 good quality and 387 Ideal quality diamonds in this sample. 

### Barcharts / Barplots
A Barchart or barplot takes these frequencies, and draws bars along the X-axis where the height of the bars is determined by the frequencies seen in the table. 

The geometry needed to draw a barchart in ggplot is `geom_bar()`.
```{r}
ggplot(dsmall, aes(x=cut)) + geom_bar()
```

## Continuous variable
The price, carat, and depth of the diamonds are all continuous variables. 

### Histograms
Rather than showing the value of each observation, we prefer to think of the value as belonging to a _bin_. The height of the bars in a histogram display the frequency of values that fall into those of those bins. 

Since the x-axis is continuous the bars touch. This is unlike the barchart that has a categorical x-axis, and vertical bars that are separated.

```{r}
ggplot(dsmall, aes(x=depth)) + geom_histogram()
```

### Density plots
To get a better idea of the true shape of the distribution we can "smooth" out the bins and create what's called a `density` plot or curve. Notice that the shape of this distribution curve is much more... "wigglier" than the histogram may have implied. 

```{r}
ggplot(dsmall, aes(x=depth)) + geom_density()
```

### Histograms + density 
Often is is more helpful to have the density (or kernal density) plot _on top of_ a histogram plot. 

* The syntax starts the same, we'll add a new geom, `geom_density` and color the line blue. 
* Then we add the histogram geom using `geom_histogram` but must specify that the y axis should be on the density, not frequency, scale. 
    - Note that this has to go inside the aestetic statement `aes()`. 
* I'm also going to get rid of the fill by using `NA` so the colored bars don't plot over the density line. 

```{r}
ggplot(dsmall, aes(x=depth)) + geom_density(col="blue") + 
  geom_histogram(aes(y=..density..), colour="black", fill=NA)
```

### Boxplots
Another very common way to visualize the distribution of a continuous variable is using a boxplot. Boxplots are useful for quickly identifying where the bulk of your data lie. R specifically draws a "modified" boxplot where values that are considered outliers are plotted as dots. 

#### base
```{r}
boxplot(dsmall$depth)
```

Notice that the only axis labeled is the y=axis. Like a dotplot the x axis, or "width", of the boxplot is meaningless here. We can make the axis more readable by flipping the plot on it's side. 
```{r}
boxplot(dsmall$depth, horizontal = TRUE, main="Distribution of diamond prices", xlab="Dollars")
```

Horizontal is a bit easier to read in my opinion. 

What about ggplot? ggplot doesn't really like to do univariate boxplots. You'll see those later when we create one boxplot per group. 

----

# Bivariate (Two Variables)

## Categorical v. Categorical

### Two-way Tables
Cross-tabs, cross-tabulations and two-way tables (all the same thing, different names) can be created by using the `table()` function. 

#### Frequency table
The frequency table is constructed using the `table()` function. 
```{r}
table(dsmall$cut, dsmall$color)
```

There are 4 Fair diamonds with color D, and 21 Ideal quality diamonds with color J.

#### Cell proportions
Wrapping `prop.table()` around a table gives you the **cell** proportions. 
```{r}
prop.table(table(dsmall$cut, dsmall$color))
```
0.4% of all diamonds are D color and Fair cut, 2.1% are J color and Ideal cut. 

#### Row proportions
To get the **row** proportions, you specify `margin=1`. The percentages now add up to 1 across the rows. 
```{r}
round(prop.table(table(dsmall$cut, dsmall$color), margin=1),3)
```

14.8% of all Fair quality diamonds are color D. 5.4% of all Ideal quality diamonds have color J.

#### Column proportions
To get the **column** proportions, you specify `margin=2`. The percentages now add up to 1 down the columns. 
```{r}
round(prop.table(table(dsmall$cut, dsmall$color), margin=2),3)
```

2.7% of all D color diamonds are of Fair quality. 44.7% of all J color diamonds are of Ideal quality. 


### Grouped bar charts
To compare proprtions of one categorical variable within the same level of another, is to use grouped barcharts. 

Plot the cut on the x axis, but then `fill` using the second categorical variable. This has the effect of visualizing the **row** percents from the table above. The percent of color, within each type of cut. 

```{r}
ggplot(dsmall, aes(x=cut, fill=color)) + geom_bar()
```

The default is a stacked barchart. So add the argument `position=dodge` inside the `geom_bar` layer to put the bars side by side. 
```{r}
ggplot(dsmall, aes(x=cut, fill=color)) + geom_bar(position = "dodge")
```

And look, an automatic legend. What if I wanted to better compare cut within color group? This is the **column** percentages. Just switch which variable is the x axis and which one is used to fill the colors!
```{r}
ggplot(dsmall, aes(x=color, fill=cut)) + geom_bar(position = "dodge")
```

And this easy change is why we love `ggplot2`. 

## Continuous v. Continuous 

### Scatterplot
The most common method of visualizing the relationship between two continuous variables is by using a scatterplot. 

With ggplot we specify both the x and y variables, and add a point. 
```{r}
ggplot(dsmall, aes(x=carat, y=price)) + geom_point()
```

### Adding lines to the scatterplots 
Two most common trend lines added to a scatterplots are the "best fit" straight line and the "lowess" smoother line. This is done by adding a `geom_smooth()` layer. 

```{r}
ggplot(dsmall, aes(x=carat, y=price)) + geom_point() + geom_smooth() 
```

Here the point-wise confidence interval for this lowess line is shown in grey. If you want to turn the confidence interval off, use `se=FALSE`. Also notice that the smoothing geom uses a different function or window than the `lowess` function used in base graphics. 

Here it is again using the `ggplot` plotting function and adding another `geom_smooth()` layer for the `lm` (linear model) line in blue, and the lowess line (by not specifying a method) in red.

```{r}
ggplot(dsmall, aes(x=carat, y=price)) + geom_point() + 
  geom_smooth(se=FALSE, method="lm", color="blue") + 
  geom_smooth(se=FALSE, color="red")
```


## Continuous v. Categorical
Create an appropriate plot for a continuous variable, and plot it for each
level of the categorical variable. 


### Grouped boxplots

A simple addition, just define your x and y accordingly. Specifying your `fill` to be the same variable as your `x`, gives you an automatic legend. 

```{r}
ggplot(dsmall, aes(x=color, y=carat, fill=color)) + geom_boxplot()
```

### Adding violins to the boxplot
A violin plot is like a density plot, turned on it's side, and reflected around the axis for symmetry purposes. Overlaying a boxplot and a violin plot serves a similar purpose to Histograms + Density plots. It shows outliers, the location of most the data, and better shows the shape/skew of the distribution

```{r}
ggplot(dsmall, aes(x=color, y=carat, fill=color)) +
        geom_violin(alpha=.1) + 
        geom_boxplot(alpha=.5, width=.2)
```

### Grouped histograms

By default ggplot wants to overlay all plots on the same grid. This doesn't look to good with histograms or density plots. The top layer hides the others. 
```{r, fig.width=10}
a <- ggplot(dsmall, aes(x=carat, fill=color)) + geom_histogram()
b <- ggplot(dsmall, aes(x=carat, fill=color)) + geom_density() 
gridExtra::grid.arrange(a,b, ncol=2)
```

The solid fills are still difficult to read, so we can either turn down the alpha (turn up the transparency) or only color the lines and not the fill. 
```{r, fig.width=10}
d <- ggplot(dsmall, aes(x=carat, fill=color)) + geom_density(alpha=.2)
e <- ggplot(dsmall, aes(x=carat, col=color)) + geom_density() 
gridExtra::grid.arrange(d,e, ncol=2)
```

These last few plots demonstrate how to place two plots side by side. 

1. Each plot is stored as an object (`a`, `b`, `d`, `e`)
2. The `grid.arrange` function is located within the `gridExtra` package, plots the specified objects in the number of columns specified by `ncol`. 

-----

# Additional Resources

For a **full** , and comprehensive reference guide on how to do nearly anything in ggplot -- this is by far my favorite reference http://www.cookbook-r.com/Graphs/ I reference things in there (like how to remove or change the title of a legend) constantly. 

* R Graphics: https://www.stat.auckland.ac.nz/~paul/RGraphics/rgraphics.html The best book about using base graphics
* STHDA: Statisical tools for high-throughput data analysis. http://www.sthda.com/english/
* Quick-R: [Basic Graphs](http://www.statmethods.net/graphs/index.html)
* Quick-R: [ggplot2](http://www.statmethods.net/advgraphs/ggplot2.html)
* Books 
    - ggplot2 http://ggplot2.org/book/ or http://amzn.com/0387981403
    - qplot http://ggplot2.org/book/qplot.pdf
* Help lists
    - ggplot2 mailing list http://groups.google.com/group/ggplot2
    - stackoverflow http://stackoverflow.com/tags/ggplot2
    - Chico R users group
    
For any other Google Search be sure to limit searches to within the past year or so. R packages get updated very frequently, and many functions change or become obsolete. 

