---
title: "Lesson 08 - Creating graphics"
author: "Robin Donatello"
date: "Last Updated `r format(Sys.Date(), '%m-%d-%Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
---


```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
opts_chunk$set(warning=FALSE, message=FALSE) 
```

> This lesson is a brief excerpt from Dr. Donatello's [[Full data viz tutorial]](full_data_viz_tutorial.html). 

### Grouped histograms

By default ggplot wants to overlay all plots on the same grid. This doesn't look to good with histograms or density plots. The top layer hides the others. 
```{r, fig.width=10}
a <- ggplot(dsmall, aes(x=carat, fill=color)) + geom_histogram()
b <- ggplot(dsmall, aes(x=carat, fill=color)) + geom_density() 
gridExtra::grid.arrange(a,b, ncol=2)
```

The solid fills are still difficult to read, so we can either turn down the alpha (turn up the transparency) or only color the lines and not the fill. 
```{r, fig.width=10}
d <- ggplot(dsmall, aes(x=carat, fill=color)) + geom_density(alpha=.2)
e <- ggplot(dsmall, aes(x=carat, col=color)) + geom_density() 
gridExtra::grid.arrange(d,e, ncol=2)
```

These last few plots demonstrate how to place two plots side by side. 

1. Each plot is stored as an object (`a`, `b`, `d`, `e`)
2. The `grid.arrange` function is located within the `gridExtra` package, plots the specified objects in the number of columns specified by `ncol`. 

-----

# Additional Resources

For a **full** , and comprehensive reference guide on how to do nearly anything in ggplot -- this is by far my favorite reference http://www.cookbook-r.com/Graphs/ I reference things in there (like how to remove or change the title of a legend) constantly. 

* R Graphics: https://www.stat.auckland.ac.nz/~paul/RGraphics/rgraphics.html The best book about using base graphics
* STHDA: Statisical tools for high-throughput data analysis. http://www.sthda.com/english/
* Quick-R: [Basic Graphs](http://www.statmethods.net/graphs/index.html)
* Quick-R: [ggplot2](http://www.statmethods.net/advgraphs/ggplot2.html)
* Books 
    - ggplot2 http://ggplot2.org/book/ or http://amzn.com/0387981403
    - qplot http://ggplot2.org/book/qplot.pdf
* Help lists
    - ggplot2 mailing list http://groups.google.com/group/ggplot2
    - stackoverflow http://stackoverflow.com/tags/ggplot2
    - Chico R users group
    
For any other Google Search be sure to limit searches to within the past year or so. R packages get updated very frequently, and many functions change or become obsolete. 

