---
title: "Lesson 10 - Creating Intermediate graphics"
output: html_document
---

# Introduction 

```{r, warning=FALSE, message=FALSE}
library(ggplot2); library(dplyr)
knitr::opts_chunk$set(warning=FALSE, message=FALSE) 

```

# Appearance enhancements

## Themes

Try several themes. 

```{r}
ggplot(email, aes(x=line_breaks, y=num_char)) + geom_point() + theme_
ggplot(email, aes(x=line_breaks, y=num_char)) + geom_point() + theme_
ggplot(email, aes(x=line_breaks, y=num_char)) + geom_point() + theme_
```


## Change legend titles 

```{r}
ggplot(email, aes(y=num_char, x=number, fill=number)) + geom_boxplot() + 
  scale_fill_discrete(name="________")
```

Here, I `col`ored the points by a discrete variable, so the layer is `scale_color_discrete()`.
```{r}
ggplot(email, aes(y=num_char, x=number, col=number)) + geom_boxplot() + 
  scale________discrete(name="Size of number")
```

## Add means to boxplots. 

```{r}
ggplot(email, aes(y=num_char, x=number, fill=______)) + geom_boxplot() + 
  scale_fill_discrete(name="Size of number") + 
  stat_summary(fun.y="mean", geom="point", size=3, pch=17,color="___")
```


# Barchart enhancements

## Numbers to the top of bars. 

The biggest addition to a barchart is the numbers on top of the bars. 
```{r}
ggplot(email, aes(x=number)) + theme_bw() + 
    geom_bar(aes(y = ..count..)) + ggtitle("Frequency of emails by size of number") + 
    geom_text(aes(y=..count.. + 50, label=..count..), stat='count', size = 5)
```

## Plotting Proportions

```{r}
library(scales)
cut.props <- table(email$number) %>% prop.table() %>% data.frame()
cut.props # what does this data look like? 

ggplot(_____, aes(x=_____, y=_____)) + 
  geom_col() + 
  ________ (labels=percent) + 
  ylab("Proportion") + 
  xlab("Size of number in email") + 
  ggtitle("Proportion of emails by number size")
```



## Cleveland Dot Plots

```{r}
ggplot(cut.props, aes(x=Var1, y=Freq)) +  
  geom_point(size = 3) + 
  _________(aes(x=Var1, xend=Var1, y=__, yend=____), color='grey50')
```

# Faceting / paneling 

```{r}
ggplot(NCbirths, aes(x=mage, fill=gender)) + geom_density(alpha=.3) + facet_wrap(~____)
```

Using the `ncol` argument in the `facet_wrap()` statement. 
```{r, fig.height=6}
ggplot(NCbirths, aes(x=mage, fill=gender)) + geom_density(alpha=.3) + ________(~____, ncol=1)
```
 

# Multiple plots per window

#### ggplot

```{r}
library(_______)
_____ <- ggplot(NCbirths, aes(x=mage, fill=whitemom)) + geom_density(alpha=.2)
_____ <- ggplot(NCbirths, aes(x=mage, col=whitemom)) + geom_density() 
grid.arrange(_____,_____, ncol=2)
```


## Paneling on two variables

```{r, fig.width=10, fig.height=5}
ggplot(NCbirths, aes(x=mage, fill=whitemom)) + geom_density(alpha=.2) + _______(marital~habit)
```

# Removing NA categories. 

```{r}
NCbirths %>% _____(mage, whitemom, marital, habit) %>% ______() %>% 
  _____(aes(x=mage, fill=whitemom)) + 
  geom_density(alpha=.2) + facet_grid(marital~habit)
```

# Changing colors

## Manual

```{r}
ggplot(email, aes(x=number, fill=number)) + geom_bar() +
    ______(values=c("red", "green", "blue"))
```

There's some spectacularly named colors as well, 
```{r}
ggplot(email, aes(x=number, fill=number)) + geom_bar() +
    scale_fill_manual(values=c("_______", "_______", "_______"))
```
 
## Using a color palette (recommended)
```{r}
library(RColorBrewer)
ggplot(email, aes(x=number, fill=number)) + geom_bar() +
    scale_fill_brewer(palette="_____")
```


