---
title: "Day 2 notes"
author: "Robin Donatello"
date: "September 4, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import a tab delimited text file
The file path "J:/MATH130" is the path on *my* computer. You need to find your own path. 

```{r}
email <- read.table("J:/MATH130/email.txt", header=TRUE, sep="\t")
str(email)
```

to show the first few rows of the data use `head()`

```{r}
head(email)
```


# reading in excel files
requires the `readxl` package. You  must load this package before you use the `read_excel` function. 

There are several ways to install packages. You only need to install a package once. 

1. type in the console `install.packages("readxl")`
2. use the package tab (lower right)
3. R studio Tools -> Install packages

```{r}
library(readxl) # loads the library/package
police  <- read_excel("J:/MATH130/WashPost Police Shootings 2015.xlsx", sheet=1, col_names=TRUE)
police[1:10,1:5]

```



# reading in CSV files
use the function `read.csv()`

```{r}
NCbirths <- read.csv("J:/MATH130/NCbirths.csv", header=TRUE)
head(NCbirths)
```

try to find the mean of the fathers age `fage`
```{r}
mean(NCbirths$fage)
```

use table to find missing values in categorical data by adding the `useNA="always"` argument. 
```{r}
table(NCbirths$habit, useNA="always")
```

without that argument, no NA's are shown

```{r}
table(NCbirths$habit)
```




