---
title: "Day3 notes"
author: "Robin Donatello - section 01"
date: "September 4, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# reading in tab delimited data
using `read.table()`. Don't forget to turn all slashes forward. 
The file path "J:/MATH130" is the path on *my* computer. You need to find your own path. 

```{r}
email <- read.table("J:/MATH130/email.txt",header=TRUE, sep="\t")
```

Then run this code chunk and look in your environment tab(top right) to confirm that the data set has 3,921 observations and 21 variables. 

ways to look at your raw data: 

1. click the blue arrow in the upper right
2. use the `str()` function

```{r}
str(email)
```

the `head()` and `tail()` functions print out the first (or last) 6 rows in the data set

```{r}
head(email)
```

you can also view the dataset in a spreadsheet like format by clicking on the spreadsheet icon to the right of the data name in your environment. 

# Excel

Same deal, read in an excel file. 
This uses the `readxl` package. (read X-L not x-one) 
If you have not installed the packaget yet, do so in 1 of these ways: 

* type in the console: `install.packages("readxl")`
* use the package tab in the lower right. click "install"

remember you only have to install packages once. 

```{r}
library(readxl) # load package / load library
police  <- read_excel("J:/MATH130/WashPost Police Shootings 2015.xlsx", sheet=1, col_names=TRUE)
police[1:10,1:5]
```


# Comma separated values
use `read.csv()`

```{r}
NCbirths <- read.csv("J:/MATH130/NCbirths.csv", header=TRUE)
head(NCbirths)
```

arithmetic functions such as the mean of a variable that contains missing data will result in a missing value. 
```{r}
mean(NCbirths$fage)
```

the `summary()` function calculates summary statistics for a numerical variable.
```{r}
summary(NCbirths$fage)
```

For categorical variables (Factors), use a frequency table. 

```{r}
table(NCbirths$habit)
```

to show if there are any  missing values, add the `useNA` argument. 
```{r}
table(NCbirths$habit, useNA="always")
```
