---
title: "Lab 2: Data import, functions and factors"
author: "NAME"
date: "DATE"
---

Answer the questions in this lab and submit the compiled HTML or PDF by the deadline. 


1. What error did I make in the second example of converting the gender variable to a factor
  using the code:  `parHIV$gf2 <- factor(parHIV$GENDER, labels=c("M", "F"))`?  
  
```{r, eval=FALSE}
parHIV$gf2 <- factor(parHIV$GENDER, labels=c("M", "F"))
table(parHIV$gf2, parHIV$GENDER)
```
  
    - How was this problem detected? 
    - How would you fix it? 
2. When reading in the `parHIV` Excel data set, what do the arguments `sheet=1` and `col_names=TRUE` mean? 


Use the `NCbirths` data set to answer the next set of questions. Read in the
data set in the code chunk below. 

```{r}
# Read in the NC Births data set here, then delete this comment
```


1. How many observations and variables are contained in this data set? 


2. Calculate the mean age of the mothers (`mage`) in the sample.


3. Calculate the trimmed mean mothers age after eliminating the outer 10% of ages


4. Use the `summary` function to calculate summary statistics on the fathers age (`fage`). 
   Round to 3 digits using the `digits=` argument. Don't forget that you can look at the 
   bottom of the help for summary (`?summary`) file for examples on how to use this function.  
   
   
5. What is the distribution of smoking habit (`habit`) amongst the mothers in the sample? 
   i.e. How many are smokers and how many are non-smokers?  _Hint: Use the `table()`_ function. 
   
   
6. Use the following code to create a new indicator variable for if the fathers age is missing: 
   `NCbirths$miss_fage <- ifelse(is.na(NCbirths$fage), 1, 0)`. What data type is this new variable?
   
   
7. Convert `miss_fage` to a factor variabale, where the lables are "Yes" when `miss_fage`=1 and 
   "No" when `miss_fage` = 0. Create a `table()` of this new variable to show that your labeling 
   worked. 
   
   
   
8. What proportion of fathers ages are missing?  




